---
title: "MARS"
author: Zvikomborero Kennedy Jokonya
format: html_document
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# MARS - Multivariate Adaptive Regression Splines. Implementation using R

---

## Overview & Description

This MARS package uses Freidman's (1991) algorithms to implement Multivariate Adaptive Regression Splines in regression models.The benefit to using MARS is its flexibility on adapting to non-linear regression models by using stepwise recursive partitioning on basis functions and the interaction with predictors. 

The goal of mars is to ...

## Files:

| File | Description |
|------|-------------|
| `mars.R` | Contains: mars(), with fwd_stepwise(), bwd_stepwise(), LOF(), h(), init_B(), split_points(). Also , Validator - validate_mars.control(),  helper - new_mars.control and constructor for mars.control(). These.......... |
| `predict.mars.R` | Contains: predict.mars(), make_B() and h(), to.......... |
| `plot.mars.R` | Contains: plot.mars() which generates plots for fitted vs actual, residuals vs fitted, a Q-Q plot, and residuals vs index............|
| `print.mars.R` | Contains: print.mars(), to display mars model.......... |
| `summary.mars.R` | Contains: summary.mars(), which uses summary() method for fitted model......... |
| `test.R` | Contains:........... |


## Dependencies

- Base R (May be up to date version)

## Installation

You can install the development version of mars like so:
``` r
# From github repository:
devtools::install_github("butterbumbs/MARS")

# From downloaded source files:
source("mars.R")
source("predict.mars.R")
source("plot.mars.R")
source("print.mars.R")
source("summary.mars.R")

```

## Use

This is a basic example which shows you how to solve a common problem:

```{r example}
library(mars) 


# Fit model
mc <- mars.control(Mmax = ...)
fit <- mars(y ~ . , data = data, control = mc)

#  methods
print(fit)
summary(fit)
plot(fit)
predict(fit, newdata = newdata)
```

## Author

| Zvikomborero Kennedy Jokonya | email: zkj@sfu.ca | github: @butterbumbs |

## License

This project is licensed under the MIT License - see the LICENSE.md file for details


## References

Jerome H. Friedman "Multivariate Adaptive Regression Splines," The Annals of Statistics, Ann. Statist. 19(1), 1-67, (March, 1991)
