---
title: "mars: Multivariate Adaptive Regression Splines"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mars}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Description

## Usage

```{r eval=FALSE}

# default settings
mars(formula, data, control = mars.control(Mmax=2,d=3,trace=FALSE))
# Example:
formula =  y ~. 
data = data
control = mars.control(Mmax=2,d=3,trace=FALSE)
# set trace=TRUE to trace output

# 
mars(y~. , data, control = mars.control())


```

## Arguments


- `formula` -
- `data` -
- `control` -

## Details

## Value

- `call` -
-`formula` -
- `y` -
- `B` -
- `Bfuncs` -
- `x_names` -
- `...` -

## Author

| Zvikomborero Kennedy Jokonya | email: zkj@sfu.ca | github: @butterbumbs |

## References

Jerome H. Friedman "Multivariate Adaptive Regression Splines," The Annals of Statistics, Ann. Statist. 19(1), 1-67, (March, 1991)

## See Also

- `predict.mars()` — 
- `print.mars()` — 
- `summary.mars()` — 
- `plot.mars()` — 
- `mars.control()` — 
- `anova.mars()` — 

## Examples

```{r}
# Example 1:

```

```{r}
# Example 2: Economic data
# Using ISLR data to find the best predictor of wage
library(mars)
set.seed(2026)

Wage <- ISLR::Wage[,-10] # drop logwage or replace wage with logwage as the response

fit <- mars(wage ~. , Wage, control = mars.control(Mmax=6))
print(fit)
summary(fit)
plot(fit)
head(predict(fit))
anova(fit)

### excluding categorical and dummy variables
fit_2 <- mars(wage ~ age + year , Wage, control = mars.control(Mmax=6))
print(fit_2)
summary(fit_2)
plot(fit_2)
head(predict(fit_2))

mod <- lm(wage ~ age + year , Wage)
summary(mod)
```

```{r echo=FALSE, message=TRUE, warning=TRUE, results='hide'}
b2 <- round(fit$coefficients[2], 2)
b2
b4 <- round(fit$coefficients[3], 2)
b4
b6 <- round(fit$coefficients[4], 2)
b6
```
The model in fit:


Using a mars model on data from ISLR, it was able to prune predictors to an Mmax=6. that resulted in four basis functions of significnce. Variables included in the final output are: education (Advanced Degree & College Grad), and health insurance (= No). So this can be interated as; an individual's wage is associated with an increse of $`r b2` if they have an advanced degree, a decrease of `r b4` if they do not have health insurance and an increase of `r b6` if they are at least a college grad too. 

The predictors choosen are categorical and dummy variables, which capture a logistic relationship. `fit_2` captures a numerical relationship which only includes age and year, where the interpetation work similarily, but shows the non linear relationship. 

```{r}
# Example 3: Astronomy

```

```{r}
# Example 4: Sound/Music

```
