---
title: "mars: Multivariate Adaptive Regression Splines"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mars}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Description

This MARS package uses Freidman’s (1991) algorithms to implement Multivariate Adaptive Regression Splines in regression models.The benefit to using MARS is its flexibility on adapting to non-linear regression models by using step wise recursive partitioning on basis functions and the interaction with predictors.

Using step-functions between knots, it splits basis functions and removes those that don't improve the model. Returning the optimal number of basis functions.

## Usage

```{r eval=FALSE}

# default settings
mars(formula, data, control = mars.control(Mmax=2,d=3,trace=FALSE))
# Example:
formula =  y ~. 
data = data
control = mars.control(Mmax=2,d=3,trace=FALSE)
# set trace=TRUE to trace output

# 
mars(y~. , data, control = mars.control())

```

## Arguments

-   `formula` - object of formula to regress a response such as y and a predictor such as x. Use . for all in data
-   `data` - input data frame
-   `control` - mars.control object - choice of Mmax

## Details

A fitted model using mars, primarily go through three stages. First when a call is made, it goes through a forward pass stepwise selection of basis functions in fwd_stepwise(), which creates splits on basis functions and optimizes these to find the best splits. Second, the results of fwd_stepwise, are passed to a backward stepwise selection of basis functions in bwd_stepwise(), which prunes basis functions by removing those that do not improve the model by checking Generalized Cross-Validation (GCV) and when GCV can no longer be reduced, we have our output. The final stage returns the results of the best basis functions, with their variable/predictor, sign and knot of their split. The output returned is augmented by additional parameters from ⁠mars.control().⁠

## Value

an object of class mars, inheriting from class lm. They return a list of class lm plus list of class 'mars.

Some `lm` values:

-   `coefficients`.. holds coefficients on model

-   `rank`.......... rank of basis function

-   `residuals`..... error terms

-   `etc`........... additional `lm` values

Additional to this list in `mars` is:

-   `formula`....... object of formula

-   `call`.......... call of formula matched

-   `y`............. data frame of response

-   `B`............. basis functions

-   `Bfuncs`........ optimize basis functions

-   `x_names`....... names of predictors

## Author

| Zvikomborero Kennedy Jokonya \| email: zkj\@sfu.ca \|

## References

Jerome H. Friedman "Multivariate Adaptive Regression Splines," The Annals of Statistics, Ann. Statist. 19(1), 1-67, (March, 1991)

## See Also

-   `predict.mars()`
-   `print.mars()`
-   `summary.mars()`
-   `plot.mars()`
-   `mars.control()`
-   `anova.mars()`

## Examples

```{r}
library(mars)
# Example 1:

m <- mars(y~., marstestdata,control = list(Mmax = 2L, d = 3, trace = FALSE))
summary(m)
print(m)
predict(m, newdata = marstestdata)
anova(m)
plot(m)
```

```{r}
#Example 2: Spotify data, predicting popularity of songs
#data from: https://www.kaggle.com/datasets/maharshipandya/-spotify-tracks-dataset

fit <- mars(popularity ~. , spotify, control = mars.control(Mmax=4))
summary(fit)
anova(fit)
```

```{r}
# Example 3: Airbnb prediction on rental price
## dataset from: https://www.kaggle.com/datasets/dgomonov/new-york-city-airbnb-open-data/

airbnb <- na.omit(airbnb)

fit <- mars(price ~. , airbnb, control = mars.control(Mmax=4))
summary(fit)
anova(fit)
plot(fit)

```

```{r}
# Example 4: Pima Indians Diabetes (Pima) dataset, predict chance of diabetes for women
# data from: https://machinelearningmastery.com/standard-machine-learning-datasets-for-imbalanced-classification/

fit <- mars(X0.627 ~. , diabetes, control = mars.control(Mmax=4))
summary(fit)
predict(fit, newdata = diabetes)
anova(fit)
plot(fit)
```

```{r}
# example 5
set.seed(2026)
x1 <- 1:100
x2 <- rnorm(100)
x3 <- rnorm(100)
y <- pmax(0, x1 - 0.5)*pmax(0, x2 - 0.5)  + rnorm(100)
data <- data.frame(y,x1,x2,x3)
fit <- mars(y ~ ., data = data, control = mars.control(Mmax=4))
plot(fit)
```